Weighing Chances and Use Against Risks and Misuse of Emerging Genome Analysis and Manipulation Technology

### 2022-05-31 draft.tex
Spelling error corrected

\begin{abstract}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is the molecule of life, DeoxyriboNucleic Acid. The genetic material that after being refined by millenia of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how applied genetics will impact us in regard to ethics and data privacy. 
Since great responsibility comes with great power, Genetic engineering is as much blessing as curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well and enables more and more accurate predictions about the future.
This paper summarises the chances and risks of genome analysis broken down in three fundamental questions. \textit{Who are you} focuses on the individual consumer of DNA tests, what is expected to gain from the results and why it is growing in popularity. \textit{What have you done} takes a look at the past of an individual and what information can be tracked by DNA. \textit{What does your future look like} aims at the consequences this information can have.
\end{abstract}
\section{Historical and Technical Introduction}

@@

\subsection{Structure of DNA}
To aid a discussion a short primer on what DNA actually is may be appropriate, the well initiated reader may skip this section.
DNA is an acidic molecule present in every complex organism, which encodes the cell function. In its natural form it takes the shape of a double helix. The outer rails, called backbones are made up of phosphate and a pentose sugar called deoxyribose. Each ladder step connecting the backbones is a pair of nitrogenous bases, also called nucleobases. These bases come in the form of chemical compounds called purine (adenine-A \& guanine-G) and pyrimidines (cytosine-C \& thymine-T) which are connected via hydrogen bond.
These bases form are complementary, meaning a purine always bonds with the corresponding pyrimidine to form the base pairs GC and AT.
DNA can be cloned by breaking the hydrogen bond, as the rules of complementarity dictate where each free floating base should attach itself.
Within a cell DNA may be read and transcribed into RNA, a similar but single stranded structure where the nitrogen base uracil-U replacess thymine. RNA is then either processed further translated into amino acids to produce proteins or be expelled from the cell\cite{dnaprimer}. This process is guided by codons, three base long combinations of which there are $4^3 = 64$.
Except for the start and three stop codons (AUG, UAA, UAG and UGA respectively), which guide where to start and stop translation, each codon codes for one of twenty amino acids\cite{codon}.
To get an idea of the factors involved in this procedure, the exact enzymes which break apart, or process DNA and its derivates are irrelevant. It is simply important to understand that every protein that makes up life is created from amino acids, the why, when, how, how much and what of which is encodable in the order of short snippets of bases that lie between start and stop codons.

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artifically pollinated date trees\cite[p.633]{mayr}.
Since then several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic matieral to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a seperate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.
The most direct and controlled form of genetic modifiction is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Secondly gene association studies, which aim to map genetic variants to observable traits and come in two varieties. 
Genome-wide association studies (GWAS) hope to uncover any correlation between strings of DNA and phenotype, while specific genetic association or gene linkage studies focus on specific variants of a gene (alleles) and examine whether they frequently cooccur with disease\cite{gwas}.
This endeavour presents numerous challenges stemming from the sheer complexity of the mechanisms at hand.
Part of the reason this differs from an exact science is changes an organisms environment has on its genome. It is estimated as much 98\% of human DNA is noncoding, id est lacks any apparent purpose\cite{noncoding}. Some of it however is not entirely disfunctional, rather it is dynamically activated and silenced in a process known as epigenetics changes by so called regulatory genes\cite{epigenetics}.
This means the exact same sequence of DNA can express itself differently, making direct conclusions about the significance of a specific snippet fall short of reliability.
In 2003 with the official completion of the Human Genome Project (HGP) this has been accomplished\cite{HGPFAQ}.
The human genome consists of approximately 3.3 billion basepairs, by cross referencing a healthy genome, single point mutations can be effortlessly identified in theory, especially with modern high-throughput analysis tools\cite{rawal}.
Unfortunately in the context of genetics this notion of a healthy genome is vague. Not only does a healthy genome not exist, simply averaging the genome of a few relatively healthy adults would do it no justice. Changes in DNA can also come in the form of deletion or addition of bases rendering traditional similarity metrics useless as an applicable method would have to be alignment free, in other words agnostic to the exact position in sequence, rather it should recognize subsequences or repetitions anywhere they show up and compare the frequency. Through the nature of mutation, the 'average genome' also varies widely regionally, the HGP in particular faced criticism for overrepresenting caucasian DNA, owed to the fact that it was conducted on anonymous donations of DNA in the USA\cite{HGPcritic}.
It should therefore be seen more like a reference frame to possibly detect deviation.
Cataloguing mutations is useful but by itself often provides little insight, as chances are the mutation occured a noncoding part.
Even when mutation occurs in regulatory sequences or as part of a gene single mutations these usually have little effect.
Its important to keep in mind that with mutations being so common, each cell has a slightly different sequence. The only mutations they share are those that occured in the germ cells that formed the zygote or in the very early stages afterwards.
Germline mutations occur roughly $\frac{number of base pairs}{2}*10^{-9}$ times per year. In humans that works out to about 1.65 inheritable mutations per year per parent, but when examining the genome of adult humans there is sampling bias at play. Single point mutations with drastic changes usually just render the cells nonviable, consequently an egg being fertilized and developing long enough to be given birth to indicates a mostly intact genome\cite{scally}.
Somatic mutations occur in the order of magnitute of $ 2*\frac{number of cell divisions}{10^8 base pairs}$, which with 30 trillion cells equates to roughly 500 trillion per year\cite{somaticmutation}.
These usually have no or very little effect, are repaired by the cells dna repair mechanisms or kill the cell either by inhibiting vital function or by causing cell suicide\cite{dnarepair}.
The unlikely unlikely case that DNA damage results in neither of these scenarios and the cell instead starts rapidly reproducing is what we call cancer.

@@

Every individual has a unique genetic makeup, their own distinct form of the human genome. 
Furthermore, our DNA remains the same from the first instant of an individual’s existence to his or her last breath.\cite{dnaidentity} 
Sociologist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term.\cite{abelsidentity} Consequently, it is not unjustified to combine both. On the one hand, there is a distinctive combination of DNA sequences that is individual for everyone and consistent for life. 
On the other hand, there is a need for clarification about one's own identity, the will to differentiate oneself from others while remaining consistent within oneself. But, what do people hope to gain from their genetic test results?

@@

The two biggest market players Ancestry and 23andMe have a combined DNA Database Size of 32+ million\cite{ancestrydatabase}\cite{about23andme}. With this amount of data and due to the sensitivity of the content, it should be handled very responsibly. But how does it look in practice?
Laestadius, Rich \& Auer investigated 30 different direct-to-consumer genetic testing websites for two aspects. 
The first aspect being transperency regarding confidentiality and privacy. As a second aspect, the use of data, professional and government bodies created guidelines to promote transparency among these companies was analysed. 
The authors concluded that although most companies met guidelines related to transparency regarding security protocols, storage procedures, and third-party disclosures, only few met guidelines regarding sharing risks from data disclosures. 
Additionally, few companies disclosed how long data would be kept for services or research. 
Use of data for research was frequently mentioned only in privacy policies and terms of service documents, and only two-thirds of companies required an additional consent to use consumer data for health-related research. 
This analysis shows that companies do not consistently meet international transparency guidelines related to confidentiality, privacy, and secondary use of data and are therefore a risk for their customers. 
The authors appeal to clearly inform consumers about specific third parties that will be given access to their data, whether they can have their samples and data deleted at will, and whether their data will be used for anything other than the services they purchased.\cite{companydata}

@@

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non negotiable, which neccessitates public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.
\subsubsection{Genetic Discrimation}
As will be discussed in the following chapter, subpar treatment grounded in genome analysis results is no better than discrimination on the basis of sex, race, age or religion.
\subsubsection{Commercial Use}
Profit as the primary incentive manages to corrupt even the most honourable endeavour.
Commercial gene testing and therapy raises the same questions about motive and ethical obligation to help that plagued the medical ethics for millenia.
Testing costs being inflated because of patents can hardly be in accordance with public interest, which sparks discussion about the next point\cite{publicopinion}.
\subsubsection{Use of Public Resources}
If private patents are to be avoided, public resources have to be used towards research in the applicable area.
Distributing those in a fair way however proves difficult. Developing tests is expensive so some conditions need to be prioritized, which in turn requires weighing of different factors. The initial cost of development is just as important as benefit it brings. This is comprised of the reliability of the result, as well as therapy options versus their respective cost. An unreliable result is equally useless from a public health perspective as knowledge about a genetic death sentence.
Additionally as many tests would only ever affect a small number of people, some argue that the money should either be used for more general genetic tests or towards programs that target public health in a broader sense like adressing diet\cite{publicopinion}.
\subsubsection{Patient Responsibilities}
While a general right to data privacy in medicine makes sense for traditional diagnostics, one might argue that since DNA test results have immediate consequences for relatives theres at least in part a responsibility to inform said relatives of any risks they themselves might face, comparable to infectious disease or STDs\cite{publicopinion}.
This apparent obligation to warn stands in direct conflict with the testers pledge of silence and the testees right to privacy, a conflict that could in theory be resolved with institutions for anonymous testing and subsequent information of concerned parties analogous to HIV testing.
\subsubsection{Psychological Risks}
The psychological weight associated with carrying near certain information about the future is not to be underestimated in the context of genetic testing either.
Just as a cancer diagnosis, discovering to be at high risk of devoloping potentially lethal disease can be devastating, especially if treatment options are limited\cite{publicopinion}.
Depending on the evolution of public debate on the topic, a scenario in which gene therapy is frowned upon in a similar manner to termination of pregnancy by groups within society is also imaginable.
This would add additional stress and force people at risk for genetic disease to make a choice between social acceptability and their own health.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
Looking at the probability theory behind it, we expect to pay more than the 
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance, except possibly against inflation.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing gene testing aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and taking precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.
It is imperative that sensitive data continues to be stored discretely access is tightly controlled as to not enable exploitation of this knowledge.
Were genetic discrimination to become a common occurence, would it be likely that the tradeoff for a chance at a healthier life would no longer be worth it to certain people, which would lead to unneccessary and preventable suffering.
It is therefore vital that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive informations is stored discretely and genetic discrimination is kept under control.

@@

Assuming policies would be enacted on a national level instead of as a global initiative, practicing eugenics via genetic manipulation would deepen the divide between developed and developing nations causing additional inequality, alongside the social inequality unavoidibly caused by liberal eugenics, the approach of leaving the use of enhancements to choice.
The opposite, obligatory genetic enhancement, would unearth even bigger concerns regarding bodily and reproductive autonomy\cite{eugenics}.

Artificially narrowing the human genepool might also have unforeseen sideeffects. As diversity is one of the driving agents of natural selection, careless intervention could inhibit our natural ability to defend against novel threats and disease\cite{mayr}.
It might quite literally be the end of evolution and the start of a new biological era.
There are also outcries from a religious perspective arguing that taking responsibility away from god is divinely forbidden\cite{eugenics}.

All of these have to be weighed against the possible good that genetic enhancement might bring and making a definitive recommendation free of moral ambiguity is most likely impossible.
It quickly becomes a discussion about what it is we as a society are trying to accomplish. For minimizing future suffering overall eugenics might be the obvious choice.
On the other hand a statement about maximizing joy cannot be made as easily. While healthier lives are certainly desirable, it would quickly become the status quo and no longer something to be grateful for.
With standards rising it could be equally difficult to fit in so theres no reason to believe a genetically enhanced society would constantly be ecstatic about its accomplishments.
Futhermore most of the prior arguments assume total confidence in the human ability to manipulate genes, which quite frankly is far from being warranted.
Theres good reason most of the biological world has agreed to put experimentation in humans on hold\cite{dontedit}.

>>>>

\begin{abstract}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is the molecule of life, DeoxyriboNucleic Acid. The genetic material that after being refined by millennia  of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how applied genetics will impact us in regard to ethics and data privacy. 
Since great responsibility comes with great power, Genetic engineering is as much blessing as curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well and enables more and more accurate predictions about the future.
This paper summarises the chances and risks of genome analysis broken down in three fundamental questions. \textit{Who are you} focuses on the individual consumer of DNA tests, what is expected to gain from the results and why it is growing in popularity. \textit{What have you done} takes a look at the past of an individual and what information can be tracked by DNA. \textit{What does your future look like} aims at the consequences this information can have.
\end{abstract}
\section{Historical and Technical Introduction}

@@

\subsection{Structure of DNA}
To aid a discussion a short primer on what DNA actually is, may be appropriate, the well initiated reader may skip this section.
DNA is an acidic molecule present in every complex organism, which encodes the cell function. In its natural form it takes the shape of a double helix. The outer rails, called backbones are made up of phosphate and a pentose sugar called deoxyribose. Each ladder step connecting the backbones is a pair of nitrogenous bases, also called nucleobases. These bases come in the form of chemical compounds called purine (adenine-A \& guanine-G) and pyrimidines (cytosine-C \& thymine-T) which are connected via hydrogen bond.
These bases form are complementary, meaning a purine always bonds with the corresponding pyrimidine to form the base pairs GC and AT.
DNA can be cloned by breaking the hydrogen bond, as the rules of complementarity dictate where each free floating base should attach itself.
Within a cell DNA may be read and transcribed into RNA, a similar but single stranded structure where the nitrogen base uracil-U replaces thymine. RNA is then either processed further translated into amino acids to produce proteins or be expelled from the cell\cite{dnaprimer}. This process is guided by codons, three base long combinations of which there are $4^3 = 64$.
Except for the start and three stop codons (AUG, UAA, UAG and UGA respectively), which guide where to start and stop translation, each codon codes for one of twenty amino acids\cite{codon}.
To get an idea of the factors involved in this procedure, the exact enzymes which break apart, or process DNA and its derivates are irrelevant. It is simply important to understand that every protein that makes up life is created from amino acids, the why, when, how, how much and what of which is encodable in the order of short snippets of bases that lie between start and stop codons.

\subsection{Genetic Engineering}
Human interference in evolution to bend nature to our will goes back at least to 2000 B.C. with strong evidence that the Assyrians artificially pollinated date trees\cite[p.633]{mayr}.
Since then, several methods with the shared goal of reinforcing desirable characteristics and eradicating undesirables ones developed, most dominantly selective breeding, with the recent modification called mutation breeding, the practice of directly exposing organisms, usually seeds, to mutagenic material to positively augment the mutation rate and such the chance of new positive traits being present\cite{mutationbreeding}.
It can be argued that since selective and mutation breeding only ever affect coming generations it can hardly be seen as genetic modification of a single organism and is better labeled as a separate method of eugenics.
From that perspective the first milestone of genetic engineering was the Avery–MacLeod–McCarty experiment in 1944, which proved that DNA was responsible for transformation in virus strains. It built upon Griffith's experiment from 1928, which showed that potent but dead virus strains can still transfer their properties to usually harmless strains, via something called the "transforming principle".
Avery, MacLeod and McCarty proved that the relevant component for this to work was DNA and not like previously assumed specific proteins, since the phenomenon persist when the solution is treated with protein dissolving substance, but seizes to persist when treated with DNA dissolving agents.
The most direct and controlled form of genetic modification is inserting strings of DNA into existing cells to transfer properties from one organism to another, deactivating- or cutting them out to suppress traits.
The most prominent example of this to date is the CRISPR/Cas method, which accomplishes this semi-reliably via the use of bacteria, but currently still struggles with off target effects, i.e. unforeseen changes outside of the intended locality.
While originally discovered in 1987, its use for genetic editing was only demonstrated in 2011 and since then a tremendous amount of research went into improving effectivity, safety and specificity, as well as finding applications in numerous in agriculture and medicine.
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
Modify a sufficient number of cells (most likely infeasible for broad changes), modify in an early stage of development, where only a number of cells exist, or germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

Secondly gene association studies, which aim to map genetic variants to observable traits and come in two varieties. 
Genome-wide association studies (GWAS) hope to uncover any correlation between strings of DNA and phenotype, while specific genetic association or gene linkage studies focus on specific variants of a gene (alleles) and examine whether they frequently cooccur with disease\cite{gwas}.
This endeavour presents numerous challenges stemming from the sheer complexity of the mechanisms at hand.
Part of the reason this differs from an exact science is changes an organisms environment has on its genome. It is estimated as much 98\% of human DNA is noncoding, id est lacks any apparent purpose\cite{noncoding}. Some of it however is not entirely dysfunctional, rather it is dynamically activated and silenced in a process known as epigenetics changes by so called regulatory genes\cite{epigenetics}.
This means the exact same sequence of DNA can express itself differently, making direct conclusions about the significance of a specific snippet fall short of reliability.
In 2003 with the official completion of the Human Genome Project (HGP) this has been accomplished\cite{HGPFAQ}.
The human genome consists of approximately 3.3 billion base pairs, by cross referencing a healthy genome, single point mutations can be effortlessly identified in theory, especially with modern high-throughput analysis tools\cite{rawal}.
Unfortunately in the context of genetics this notion of a healthy genome is vague. Not only does a healthy genome not exist, simply averaging the genome of a few relatively healthy adults would do it no justice. Changes in DNA can also come in the form of deletion or addition of bases rendering traditional similarity metrics useless as an applicable method would have to be alignment free, in other words agnostic to the exact position in sequence, rather it should recognize subsequences or repetitions anywhere they show up and compare the frequency. Through the nature of mutation, the 'average genome' also varies widely regionally, the HGP in particular faced criticism for overrepresenting caucasian DNA, owed to the fact that it was conducted on anonymous donations of DNA in the USA\cite{HGPcritic}.
It should therefore be seen more like a reference frame to possibly detect deviation.
Cataloguing mutations is useful but by itself often provides little insight, as chances are the mutation occurred a noncoding part.
Even when mutation occurs in regulatory sequences or as part of a gene single mutations these usually have little effect.
Its important to keep in mind that with mutations being so common, each cell has a slightly different sequence. The only mutations they share are those that occurred in the germ cells that formed the zygote or in the very early stages afterwards.
Germline mutations occur roughly $\frac{number of base pairs}{2}*10^{-9}$ times per year. In humans that works out to about 1.65 inheritable mutations per year per parent, but when examining the genome of adult humans there is sampling bias at play. Single point mutations with drastic changes usually just render the cells nonviable, consequently an egg being fertilized and developing long enough to be given birth to indicates a mostly intact genome\cite{scally}.
Somatic mutations occur in the order of magnitude of $ 2*\frac{number of cell divisions}{10^8 base pairs}$, which with 30 trillion cells equates to roughly 500 trillion per year\cite{somaticmutation}.
These usually have no or very little effect, are repaired by the cells dna repair mechanisms or kill the cell either by inhibiting vital function or by causing cell suicide\cite{dnarepair}.
The unlikely unlikely case that DNA damage results in neither of these scenarios and the cell instead starts rapidly reproducing is what we call cancer.

@@

Every individual has a unique genetic makeup, their own distinct form of the human genome. 
Furthermore, our DNA remains the same from the first instant of an individual’s existence to his or her last breath.\cite{dnaidentity} 
Sociologist Abels describes identity as the answer to the question \textit{who am I} and therefore, identity as a differentiating term.\cite{abelsidentity} Consequently, it is not unjustified to combine both. On the one hand, there is a distinctive combination of DNA sequences that is individual for everyone and consistent for life. 
On the other hand, there is a need for clarification about one's own identity, the will to differentiate oneself from others while remaining consistent within oneself. But what do people hope to gain from their genetic test results?

@@

The two biggest market players Ancestry and 23andMe have a combined DNA Database Size of 32+ million\cite{ancestrydatabase}\cite{about23andme}. With this amount of data and due to the sensitivity of the content, it should be handled very responsibly. But how does it look in practice?
Laestadius, Rich \& Auer investigated 30 different direct-to-consumer genetic testing websites for two aspects. 
The first aspect being transparency regarding confidentiality and privacy. As a second aspect, the use of data, professional and government bodies created guidelines to promote transparency among these companies was analysed. 
The authors concluded that although most companies met guidelines related to transparency regarding security protocols, storage procedures, and third-party disclosures, only few met guidelines regarding sharing risks from data disclosures. 
Additionally, few companies disclosed how long data would be kept for services or research. 
Use of data for research was frequently mentioned only in privacy policies and terms of service documents, and only two-thirds of companies required an additional consent to use consumer data for health-related research. 
This analysis shows that companies do not consistently meet international transparency guidelines related to confidentiality, privacy, and secondary use of data and are therefore a risk for their customers. 
The authors appeal to clearly inform consumers about specific third parties that will be given access to their data, whether they can have their samples and data deleted at will, and whether their data will be used for anything other than the services they purchased.\cite{companydata}

@@

\subsubsection{Patient Autonomy}
The primary role of any new medical instrument should be to serve, therefore patient autonomy and informed consent are non-negotiable, which necessitates  public education.
Broad, as well as targeted at populations most susceptible to specific genetic conditions.
Equally as important is access to testing facilities themselves as well as short wait times for both test and treatment\cite{publicopinion}.
\subsubsection{Genetic Discrimation}
As will be discussed in the following chapter, subpar treatment grounded in genome analysis results is no better than discrimination on the basis of sex, race, age or religion.
\subsubsection{Commercial Use}
Profit as the primary incentive manages to corrupt even the most honourable endeavour.
Commercial gene testing and therapy raises the same questions about motive and ethical obligation to help that plagued the medical ethics for millennia.
Testing costs being inflated because of patents can hardly be in accordance with public interest, which sparks discussion about the next point\cite{publicopinion}.
\subsubsection{Use of Public Resources}
If private patents are to be avoided, public resources have to be used towards research in the applicable area.
Distributing those in a fair way however proves difficult. Developing tests is expensive so some conditions need to be prioritized, which in turn requires weighing of different factors. The initial cost of development is just as important as benefit it brings. This is comprised of the reliability of the result, as well as therapy options versus their respective cost. An unreliable result is equally useless from a public health perspective as knowledge about a genetic death sentence.
Additionally as many tests would only ever affect a small number of people, some argue that the money should either be used for more general genetic tests or towards programs that target public health in a broader sense like addressing diet\cite{publicopinion}.
\subsubsection{Patient Responsibilities}
While a general right to data privacy in medicine makes sense for traditional diagnostics, one might argue that since DNA test results have immediate consequences for relatives theres at least in part a responsibility to inform said relatives of any risks they themselves might face, comparable to infectious disease or STDs\cite{publicopinion}.
This apparent obligation to warn stands in direct conflict with the testers pledge of silence and the testees right to privacy, a conflict that could in theory be resolved with institutions for anonymous testing and subsequent information of concerned parties analogous to HIV testing.
\subsubsection{Psychological Risks}
The psychological weight associated with carrying near certain information about the future is not to be underestimated in the context of genetic testing either.
Just as a cancer diagnosis, discovering to be at high risk of developing potentially lethal disease can be devastating, especially if treatment options are limited\cite{publicopinion}.
Depending on the evolution of public debate on the topic, a scenario in which gene therapy is frowned upon in a similar manner to termination of pregnancy by groups within society is also imaginable.
This would add additional stress and force people at risk for genetic disease to make a choice between social acceptability and their own health.

\subsection{Insurance}
Insurance as a business walks a fine line between solidarity and preying on the vulnerable. Risk and how we manage it as a society is studied under the name of insurance ethics, but to date the literature is sparse.
While public insurance in practice is a form of wealth redistribution, private insurance is in essence paying for peace of mind\cite{insuranceethics}.
Looking at the probability theory behind it, we expect to pay more than the 
However this social safety net only functions as long as there is uncertainty, in fact it functions better the more uncertainty there is.
This becomes obvious in consideration of the idea that clarity about the future replaces the insurance sector.
In a fully deterministic world, omniscient private insurance companies would never lose money on a customer, as premiums could be precisely the cost said customer will produce over their contract, plus a small markup; in other words, not providing any insurance, except possibly against inflation.
Keeping this in mind it seems logical that it's in the best interest of society at large to keep insurance companies mostly in the dark.
At least in the case of health insurance, this darkness is threatened by introducing gene testing aided risk analysis.
The inclusion of factors like genetic predisposition to disease in risk analysis is an incredibly helpful tool to doctors when choosing appropriate treatment and taking precautions, but becomes detrimental when accessed by insurance companies\cite{insurancegenes}.
Legal action to protect against this threat is already being taken, with the USA introducing the Genetic Information Nondiscrimination Act in 2008\cite{insurancegenes}.
It is imperative that sensitive data continues to be stored discretely access is tightly controlled as to not enable exploitation of this knowledge.
Were genetic discrimination to become a common occurrence, would it be likely that the trade-off for a chance at a healthier life would no longer be worth it to certain people, which would lead to unnecessary and preventable suffering.
It is therefore vital that we retain trust in and refine the methodology of genetic testing, while at the same time making sure the sensitive information is stored discretely and genetic discrimination is kept under control.

@@

Assuming policies would be enacted on a national level instead of as a global initiative, practicing eugenics via genetic manipulation would deepen the divide between developed and developing nations causing additional inequality, alongside the social inequality unavoidably caused by liberal eugenics, the approach of leaving the use of enhancements to choose.
The opposite, obligatory genetic enhancement, would unearth even bigger concerns regarding bodily and reproductive autonomy\cite{eugenics}.

Artificially narrowing the human genepool might also have unforeseen side-effects. As diversity is one of the driving agents of natural selection, careless intervention could inhibit our natural ability to defend against novel threats and disease\cite{mayr}.
It might quite literally be the end of evolution and the start of a new biological era.
There are also outcries from a religious perspective arguing that taking responsibility away from god is divinely forbidden\cite{eugenics}.

All of these have to be weighed against the possible good that genetic enhancement might bring and making a definitive recommendation free of moral ambiguity is most likely impossible.
It quickly becomes a discussion about what it is we as a society are trying to accomplish. For minimizing future suffering overall eugenics might be the obvious choice.
On the other hand, a statement about maximizing joy cannot be made as easily. While healthier lives are certainly desirable, it would quickly become the status quo and no longer something to be grateful for.
With standards rising it could be equally difficult to fit in so there is no reason to believe a genetically enhanced society would constantly be ecstatic about its accomplishments.
Futhermore most of the prior arguments assume total confidence in the human ability to manipulate genes, which quite frankly is far from being warranted.
There is good reason most of the biological world has agreed to put experimentation in humans on hold\cite{dontedit}.
